Enzo+Moray: Radiation Hydrodynamics Adaptive Mesh Refinement Simulations with Adaptive Ray Tracing

### 2010-08-19 ms.tex
Added timestep tests and figures.  Increased linewidth in dx-test figure.

\subsection{Test 5: Classical \ion{H}{2} region expansion}
\label{sec:test5}

@@

\subsection{Test 6: \ion{H}{2} region expansion in an isothermal
  sphere}

@@

\subsection{Test 7: Photo-evaporation of a dense clump}

@@

\subsection{Test 8: Champange flow from a dense clump}

@@

\subsection{Test 9: Irradiated Rayleigh-Taylor instability}

@@

\subsection{Test 10: Photo-evaporation of a blastwave}

@@

\subsection{Test 11: Collimated radiation from a dense clump}

@@

% \subsection{Test 12: \ion{H}{2} region in a rotating medium}
% 
% Angular momentum is prevalent throughout all astrophysical systems.
% We construct a problem with an \ion{H}{2} region that is completely
% contained within a sphere with solid-body rotation.  Self-gravity is
% considered in this test.  The sphere initially has $R = 40$ pc with a
% uniform density $\rho = 200~\cubecm$, temperature $T = 50$ K, and
% electron fraction $x_e = 10^{-6}$.  The sphere is partially
% rotationally supported by a factor of 0.25, i.e. the tangential
% velocity $v_\theta(r) = r\sqrt{GM_{\rm tot}/R}/4$, where $M_{\rm tot}
% = 1.3 \times 10^6 \Ms$.  The sphere and radiation source are located
% in the center of the domain.  The sphere is in pressure equilibrium
% with the ambient medium that has $\rho_0 = 1 \cubecm$, $T = 10^4$ K,
% and $x_e = 0.99$.  The radiation source has an ionizing luminosity $L
% = 10^{50}$ ph s$^{-1}$ with a 17.6 eV mono-chromatic spectrum.  The
% test uses AMR with a base grid of 64$^3$ and 2 levels of refinement,
% using the same density criterion as in Test 8.

% \subsection{Test 12: \ion{H}{2} region in a static potential}

@@

\subsection{Test 12: Time variations of the source luminosity}

@@

\subsection{Test 14: \ion{H}{2} region with MHD}

@@

In practice, we find good agreement within 5--10\% of analytical
solutions in Tests 1, 2, and 5 with only one energy group.  We use
Test 5 (\S\ref{sec:test5}; an expanding \ion{H}{2} region with
hydrodynamics) to probe any differences in the solution when varying
the resolution of the spectrum.  In RT09, ZEUS-MP was used to
demonstrate the effect of a multi-frequency spectrum on the dynamics
of the ionization front in this test.  Instead of a single shock seen
in the mono-chromatic spectrum, a the shock obtained a double-peaked
structure in density and radial velocity.  We rerun Test 5 with a
$T=10^5$ K blackbody spectrum sampled by 1, 2, 4, 8, and 16 frequency
bins.

@@

The previous three dependencies did not affect the propagation of the
ionization front greatly.  However in our and others' past experience
\citep[e.g.][\bf{+others!}]{Petkova09}, the timestep, especially too
small of one, can drastically affect the ionization front velocity.
Here we use Test 1 but with $64^3$ resolution to compare different
timestepping methods --- restricted changes in \ion{H}{2}
(\S\ref{sec:dt_hi}), constant timesteps (\S\ref{sec:dt_const}), and
based on incident radiation (\S\ref{sec:dt_tau}).

>>>>

\subsection{Test 5. Classical \ion{H}{2} region expansion}
\label{sec:test5}

@@

\subsection{Test 6. \ion{H}{2} region expansion in an isothermal
  sphere}

@@

\subsection{Test 7. Photo-evaporation of a dense clump}

@@

\subsection{Test 8. Champange flow from a dense clump}

@@

\subsection{Test 9. Irradiated Rayleigh-Taylor instability}

@@

\subsection{Test 10. Photo-evaporation of a blastwave}

@@

\subsection{Test 11. Collimated radiation from a dense clump}

@@

% \subsection{Test 12. \ion{H}{2} region in a rotating medium}
% 
% Angular momentum is prevalent throughout all astrophysical systems.
% We construct a problem with an \ion{H}{2} region that is completely
% contained within a sphere with solid-body rotation.  Self-gravity is
% considered in this test.  The sphere initially has $R = 40$ pc with a
% uniform density $\rho = 200~\cubecm$, temperature $T = 50$ K, and
% electron fraction $x_e = 10^{-6}$.  The sphere is partially
% rotationally supported by a factor of 0.25, i.e. the tangential
% velocity $v_\theta(r) = r\sqrt{GM_{\rm tot}/R}/4$, where $M_{\rm tot}
% = 1.3 \times 10^6 \Ms$.  The sphere and radiation source are located
% in the center of the domain.  The sphere is in pressure equilibrium
% with the ambient medium that has $\rho_0 = 1 \cubecm$, $T = 10^4$ K,
% and $x_e = 0.99$.  The radiation source has an ionizing luminosity $L
% = 10^{50}$ ph s$^{-1}$ with a 17.6 eV mono-chromatic spectrum.  The
% test uses AMR with a base grid of 64$^3$ and 2 levels of refinement,
% using the same density criterion as in Test 8.

% \subsection{Test 12. \ion{H}{2} region in a static potential}

@@

\subsection{Test 12. Time variations of the source luminosity}

@@

\subsection{Test 14. \ion{H}{2} region with MHD}

@@

The ionization front radius is within 5--10\% of analytical solutions
in Tests 1, 2, and 5 with only one energy group; however a
multi-frequency spectrum can create differences in the reactive
flows.  We use Test 5 (\S\ref{sec:test5}; an expanding \ion{H}{2}
region with hydrodynamics) to probe any differences in the solution
when varying the resolution of the spectrum.  In RT09, ZEUS-MP was
used to demonstrate the effect of a multi-frequency spectrum on the
dynamics of the ionization front in this test.  Instead of a single
shock seen in the mono-chromatic spectrum, a the shock obtained a
double-peaked structure in density and radial velocity.  We rerun Test
5 with a $T=10^5$ K blackbody spectrum sampled by $n_\nu = 1$, 2, 4,
8, and 16 frequency bins.  We use the following energies:
%
\begin{itemize}
\item $n_\nu = 1$: Mean energy of 29.6 eV
\item $n_\nu = 2$: Mean energies in bins 13.6--30 and $>$30 eV---21.1,
  43.0 eV
\item $n_\nu = 4$: Mean energies in bins 13.6--20, 20--30, 30--40, and
  $>$40 eV---16.7, 24.6, 34.5, 52.1 eV
\item $n_\nu = 8, 16$: Logarithmically spaced between 13.6 and 50 eV
  for the first $n_\nu-1$ bins, and the last bin is the mean energy
  above 50 eV.
\end{itemize}

Figure \ref{fig:nu_dep} shows the radial profiles of density,
temperature, ionized fraction, and radial velocity at $t = 200$ Myr
and $t = 500$ Myr.  All of the runs with $n_\nu > 1$ show the double
peaked features in density and radial velocities.  The mono-chromatic
spectrum misses this feature completely because all of the radiation
is absorbed at a characteristic column density.  In the
multi-frequency spectra, the higher energy photons are absorbed at
larger column densities and photo-heated this gas.  This heated gas
creates a photo-evaporative flow that collides with the innermost
shock, forming the double peaked density profile.  The $n_\nu \ge 4$
runs are indistinguishable, and the $n_\nu = 2$ spectrum only leads to
a marginally higher density in the outer shock and lower ionized
fractions and tempeatures in the ambient medium.  In effect, a
mono-chromatic spectrum can be sufficient if the problem focuses on
large-scale quantities, e.g. ionized filling fractions in reionization
calculations.  Conversely these effects may be important when studying
small-scale, e.g. photo-evaporation.

@@

\begin{figure}[t]
  \epsscale{0.75}
  \plotone{fig/ifront_dt}
  \caption{\label{fig:dt_dep1} Growth of the ionization front radius,
    compared to the analytical radius, in Test 1 with varying
    radiative transfer timesteps.  The $dn_{\rm H}/dt$ and $dI/dt$
    based timesteps provide the best accuracy, combined with
    computational efficiency because they take short timesteps when
    the \ion{H}{2} is expanding rapidly but take long timesteps when
    the photon gradients are small when $r_{\rm IF}$ is large.  At the
    final time, all but the $t = 5$ Myr constant timestep produce
    identical ionization front radii.}
  \epsscale{1}
\end{figure}
\begin{figure}[t]
  \epsscale{0.75}
  \plotone{fig/vary_dt}
  \caption{\label{fig:dt_dep2} Variable timestepping for the methods
    that limit change in neutral fraction (solid) and specific
    intensity (dotted).  The horizontal lines show the constant
    timesteps that were used in the tests.  The crossing time of a
    mean free path by the ionization front is plotted for reference.}
  \epsscale{1}
\end{figure}

The previous three dependencies did not affect the propagation of the
ionization front greatly.  However in our and others' past experience
\citep[e.g.]{Shapiro04, Mellema06, Petkova09}, the timestep,
especially too small of one, can drastically underestimate the
ionization front velocity.  Here we use Test 1 but with $64^3$
resolution to compare different timestepping methods --- restricted
changes in \ion{H}{2} ($dn_{\rm H}/dt$ based; \S\ref{sec:dt_hi}),
constant timesteps (\S\ref{sec:dt_const}), and based on incident
radiation ($dI/dt$ based; \S\ref{sec:dt_tau}).  The growth of the
ionization front radius is shown in Figure \ref{fig:dt_dep1}.  Both
the \ion{H}{2} restricted and incident radiation variable timestepping
methods agree within a few percent throughout the entire simulation,
as does the run with constant $dt = 0.1$ Myr timesteps.  With the
larger constant timesteps, the numerical solution lags behind the
analytical one, but they converge to an accurate \ion{H}{2} radii at
late times.  Even $dt = 5$ Myr timestep, which underestimated it by
35\% at 50 Myr, is within a percent of the analytical solution.

The larger constant timesteps deviate from these more accurate
solutions at early times because the photon energy gradient is large,
and thus so is the ionization front velocity.  To understand this, the
ionization front can be considered static in a given timestep.  Here
the ionizing radiation can only penetrate into the neutral gas by
roughly a photon mean path $\lambda_{\rm mfp}$.  Only in the next
timestep, the ionization front can advance.  If the timestep is larger
than $\lambda_{\rm mfp} / v_{\rm IF, anyl}$, then the numerical
solution may fall behind.

The variable timestepping of the $dn_{\rm H}/dt$ and $dI/dt$ methods
adjust accordingly to the physical situation, as seen in the plot of
timestep versus time (Figure \ref{fig:dt_dep2}).  They provide high
accuracy when the source first starts to shine.  At later times, the
ionization front slows as it approaches the Str\"{o}mgren radius, and
large timesteps are no longer necessary.  The $dI/dt$ method has a
similar timestep as the $dn_{\rm H}/dt$ method.  It is larger by a
factor of $\sim 2$ because of our choice of the safety factor $C_{\rm
  RT, cfl} = 0.5$.  This causes its calculated radius to be smaller by
1\% at $t < t_{\rm rec}$, which is still in good agreement with the
analytical value.
